---
title: "Rethinking Chapter 2"
author: "Gregor Mathes"
date: "2020-26-07"
output:
  html_document: 
    toc: true
    toc_float: true
    number_sections: true  
    theme: journal
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
```

# Introduction 

This is the first part of a series where I work through the practice questions of the second edition of Richard McElreaths [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/). Each post covers a new chapter. There are already some awesome sources for this book online like [Jeffrey Girard](https://jmgirard.com/statistical-rethinking-ch2/) working through the exercises of the first edition, or [Solomon Kurz](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/) leading through each example of the book with the *brms* and the *tidyverse* packages. However, so far I couldn't find a source providing solutions for the practice questions of the second edition in a tidy(-verse) way. My aim here is therefore to provide solutions for each practice question of the second edition, using the *tidyverse* and the *rethinking* packages. The first part of the series will cover chapter 2. 



# Easy practices

## Question 2E1

**Which of the expressions below correspond to the statement: the probability of rain on Monday?**

1. Pr(rain)
2. Pr(rain|Monday)
3. Pr(Monday|rain)
4. Pr(rain,Monday)/Pr(Monday)

Let's rephrase each expression as a statement:
1 is the probability of rain.  
2 is the probability of rain, given that it is Monday.
3 is the probability of it being Monday, given rain.  
4 is the probability of rain and it being Monday, given that it is Monday.  

It is quite obvious that 2 is the correct answer here, but 4 is correct as well. Using the joint probability definition on page 37, we can see that *PR(rain, Mondy)/Pr(Monday) = Pr(rain|Monday)*.

## Question 2E2

**Which of the following statements corresponds to the expression: Pr(Monday|rain)?**

1. The probability of rain on Monday.
2. The probability of rain, given that it is Monday.
3. The probability that it is Monday, given that it is raining.
4. The probability that it is Monday and that it is raining.

Now we have to convert each statement to an expression:
1 would be Pr(rain|Monday).  
2 would be Pr(rain|Monday).  
3 would be Pr(Monday|rain).  
4 would be Pr(Monday,rain).  

The correct answer is 3. 

## Question 2E3

**Which of the expressions below correspond to the statement: *the probability that it is Monday, given that it is raining?* **

Pr(Monday|rain)
Pr(rain|Monday)
Pr(rain|Monday)Pr(Monday)
Pr(rain|Monday)Pr(Monday)/Pr(rain)
Pr(Monday|rain)Pr(rain)/Pr(Monday)

Let’s convert each expression into a statement:
1 would be the probability that it is Monday, given that it is raining.
2 would be the probability of rain, given that it is Monday.
3 needs to be converted using the formula on page 37:
*Pr(rain|Monday)Pr(Monday) = Pr(rain,Monday)*  
4 is the same as the previous option but with division added:
*Pr(rain|Monday)Pr(Monday)/Pr(rain) = Pr(rain,Monday)/Pr(rain)*
Simplify this using joint probability:
*Pr(rain,Monday)/Pr(rain) = Pr(Monday|rain)*  
5 is the same as 4 but with the terms exchanged. So it can be interpreted (repeating all the previous work) as the probability of rain, given that it is Monday.

The correct answers are thus 1 and 4.

## Question 2E4

**The Bayesian statistician Bruno de Finetti (1906-1985) began his book on probability theory with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original, so I imagine de Finetti wanted us to shout the statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say “the probability of water is 0.7”?**

Solution copied over from [Jeffrey Girard](https://jmgirard.com/statistical-rethinking-ch2/):  
From the Bayesian perspective, there is one true value of a parameter at any given time and thus there is no uncertainty and no probability in “objective reality.” It is only from the perspective of an observer with limited knowledge of this true value that uncertainty exists and that probability is a useful device. So the statement, “the probability of water is 0.7” means that, given our limited knowledge, our estimate of this parameter’s value is 0.7 (but it has some single true value independent of our uncertainty).  


# Medium practices

## Question 2M1

**Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.**

1. W,W,W
2. W,W,W,L
3. L,W,W,L,W,W,W

The solution for this is given on page 40. Let's transform it to a pipe. As we apply three inputs to it, we can save some time by making a function of w for water and l for land. 

```{r question 2m1}
grid_approx <- function(w, l){
tibble(
  # define grid
  p.grid = seq(from = 0, to = 1, length.out = 20), 
  # define prior
  prior = rep(1, 20)
  ) %>% 
  # compute likelihood at each value in grid for 
  mutate(likelihood = dbinom(w, size = w + l, prob = p.grid)) %>% 
  # compute product of likelihood and prior
  mutate(unstd.likelihood = likelihood * prior) %>% 
  # standardise the posterior so it sums to 1
  mutate(posterior = unstd.likelihood / sum(unstd.likelihood)) %>% 
  # directly pipe that to a ggplot
  ggplot() +
  geom_line(aes(p.grid, posterior)) +
  labs(y = "posterior probability", x = "probability of water") +
  theme_minimal()
}
```

Now we can approximate the posterior for each option.  
Option 1:
```{r 2M1 option 1, fig.dim=c(7,4)}
grid_approx(w = 3, l = 0) # 3 water, 0 land
```

Option 2:
```{r 2M1 option 2, fig.dim=c(7,4)}
grid_approx(w = 3, l = 1) # 3 water, 1 land
```

Option 3:
```{r 2M1 option 3, fig.dim=c(7,4)}
grid_approx(w = 5, l = 2) # 5 water, 2 land
```
